---
title: "Reproducible Research, assignment 1"
author: "Phillip R. Kyle"
output: html_document
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
setwd("C:/Users/Phillip/Desktop/R/Chapter5/Assignment1/")
``` 
  
### Loading and preprocessing the data
***

First the activity data was downloaded.  

```{r download_data, echo=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "activity.zip")

```

Data was then loaded into the workspace and the date variable was converted to an actual date variable.

```{r}

a <- read.csv(unz("activity.zip", "activity.csv"), na.strings = "NA")
a$date <- as.Date(a$date)
```
</br>
  
### What is mean total number of steps taken per day?
***
The total number of steps per day were first calculated.  

```{r}
totalStepsDay <- aggregate(a$steps, by=list(a$date), FUN=sum,na.rm=TRUE)
names(totalStepsDay) <- c("date", "steps")
```
  
The histogram was then be produced and mean/median reported.

```{r}
hist(totalStepsDay$steps, 
     breaks = seq(0,25000, by = 1000), 
     col = "grey", 
     main = "Total number of steps per day", 
     xlab = "Steps per day",
     ylab = "Count"
     )

summary(totalStepsDay$steps, na.rm = TRUE)
```
</br>

### What is the average daily activity pattern?
***
First the average number of steps per interval across all days was calculated. 

```{r}
meanStepsDay <- aggregate(a$step, by = list(a$interval), FUN = mean, na.rm = TRUE)
names(meanStepsDay) <- c("interval", "steps")
```

The time-series plot could then produced.

```{r ggplot, include=FALSE}
require(ggplot2)
```

```{r}
#Generate time-series plot
g <- ggplot(meanStepsDay, aes(interval, steps, group = 1))
g+geom_line(color="blue")+
        scale_x_continuous(breaks = seq(0, 2400, by = 200))+ 
        labs(x ="Interval",y = "Avg. no. of steps taken", title = "Avg. number of steps taken by 5-minute intervals")
```

And the 5-minute interval with the highest avg. steps per day found. 
```{r}
meanStepsDay[order(meanStepsDay$steps, decreasing = TRUE),][1,]
```
</br>

### Imputing missing values
***
The number of missing values was first determined.

```{r}
sum(is.na(a))
```

Interval means to used to replace missing values.

```{r}
a$intervalMean[match(a$interval, meanStepsDay$interval)] <- meanStepsDay$steps
```

A new dataset was then created with the missing data filled in 
```{r}
aImp <- transform(a, steps = ifelse(is.na(steps), intervalMean, steps))
```

Using this dataset a new histogram was produced and mean/median calculated. 

```{r}
#Calculate total steps per day with missing values imputed
totalStepsImp <- with(a, aggregate(steps, by = list(date), FUN = sum))
names(totalStepsImp) <- c("date", "steps")

#Generate histogram 
hist(totalStepsImp$steps,
     breaks = seq(0,25000, by = 1000), 
     col = "grey", 
     main = "Total number of steps per day (imputed)", 
     xlab = "Steps per day",
     ylab = "Count"
     )

summary(totalStepsImp$steps, na.rm = TRUE)
     
```
</br>

### Are there differences in activity patterns between weekdays and weekends?
***

A new factor was created with two levels - "weekday" and "weekend" 

```{r}
#Assign weekday based on date
a$weekday <- tolower(weekdays(a$date))

#Determine type of day
a$daytype <- as.factor(with(a, ifelse(weekday == "saturday" | weekday =="sunday","weekend", "weekday")))
```

The data frame was split by day type, and interval means were calculated for both.

```{r}
#Split data frame to separate date-frames by day type
x <- split.data.frame(a, a$daytype)
list2env(x, envir = .GlobalEnv)

#Calculate mean by interval for weekdays
weekdayInt <- aggregate(weekday$step, by = list(weekday$interval), FUN = mean, na.rm = TRUE)
weekdayInt$daytype <- rep("weekday",length(weekdayInt$x))
names(weekdayInt) <- c("interval", "steps", "daytype")

#Calculate mean by interval for weekends
weekendInt <- aggregate(weekend$step, by = list(weekend$interval), FUN = mean, na.rm = TRUE)
weekendInt$daytype <- rep("weekend",length(weekendInt$x))
names(weekendInt) <- c("interval", "steps", "daytype")

#Bind data frames
y <- rbind(weekdayInt, weekendInt)

```


Finally, the panel plot was generated using the GGplot2 package. 

```{r}
g <- ggplot(y, aes(interval, steps, color = daytype))
g+geom_line()+facet_grid(daytype~.)+scale_x_continuous(breaks = seq(0, 2400, by = 200))+labs(x = "5-minute interval", y = "Steps")+theme_bw()
````

